AMBIENT; an execution and an MPI-CUDA scheduler

### 2011-08-05 doc/paper_ambient/paper_ambient.tex
paper ambient, packet manager, figure, plot, + first modif from Alex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN PACKAGE
\documentclass[11pt]{amsart}       % AMS template just for the draft
\usepackage{geometry}                 % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                        % .letter or a4paper or a5paper or ... 
%\geometry{landscape}                 % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}     % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}                   % FIgure
\usepackage{amssymb}                 % Math symbol
\usepackage{epstopdf}                   % for working in .pdf and  not .dvi (convert .eps, .jpg,  etc to .pdf)
\usepackage{psfrag}
\usepackage[T1]{fontenc}
\usepackage{aeguill}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKAGE

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS
\title{AMBIENT; an  execution and an MPI-CUDA scheduler  }
\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauerb, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
%\author[UNIGE, ETHZ, CSCS]{UNIGE, ETHZ, CSCS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS

@@

\section{Design of the software layer}
In this section, we describe our framework in simple terms, without  a deep technical description.
We will focus on the main functionalities  and show the essential features.

@@

Ambient employs an extension on the CUDA approach to a distributed cluster. This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which an runs an independent thread. In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc must be considered. A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.

First, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$.
Of course, the user has full control of the size and can adapt it to the treated problem. Like in CUDA, blocks (i.e. groups of "threads") and a grid of blocks can then be formed in Ambient.

@@

\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Determination of the directed  acyclic graph determination   &    \\ \hline
                           \textbf{Require}    A class Matrix  that fits with Ambient framework   &  \\  
                           and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
                           \begin{tabular}{c l} 
                               \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j$  to the size of the vector \textbf{do} \\                               
                               \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
                               \tiny{4:}   & \hspace{0.8 cm}  $A[i]+ \!\! =B[j]$, $\vartriangleright$  push l/c kernel 1, \\
                               \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =3$,  $\vartriangleright$  push l/c kernel 2, \\                   
                               \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
                               \tiny{8:}   & \hspace{0.8 cm} $C[i]\times \!\! =D[j]$,  $\vartriangleright$ push l/c kernel 3, \\
                               \tiny{9:}   & \hspace{0.4 cm}  \textbf{end if} \\       
                               \tiny{10:} & \hspace{0.2 cm}  \textbf{end do} \\   
                               \tiny{11:} & \textbf{end do} \\                                 
                               \tiny{12:} & \texttt{playout()} $\vartriangleright$  execute the stack \\
                               \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
                                                & that push the good logistics and computational kernel. \\                               
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements. The association of a cumulative stack and  delate stack execution will determine the  directed  acyclic graph. \label{GRAPH}}
\end{table}

@@

When we execute a computational kernel, we get the data by the Ambient command  \texttt{current(\textit{A})(i,j)} (ref. \ref{CURRENT}), where \textit{A} is a matrix 
compatible with Ambient, and the indices $i$ and $j$ are the coordinates of the work group (Fig. \ref{MPIGROUP}). Three possibilities are possibles for getting the data when a process execute this functions
in a computational kernel.
\begin{enumerate}
\item The work group is already in the memory of the process. The access is direct.
\item The work group is in the memory space of the present MPI-group (previously created by the associated logistics kernel). The applicant process sends a query request (a specific packet) to its  master process of its MPI group.
The master process of the group gets the request (the master process knows the repartition of the workgroup inside the MPI-group), it sends a new request to
the actual owner, when the actual owner gets the request, it send the workgroup needed to the first applicant. 
\item The work group is inside an other MPI-group. As previously, the applicant process sends to the master process a query  request, and the master process of 
the actual MPI group  sends an other request to the older owner group. The older master process gets the message and sends a message to actual owner of the data. Then the owner 
of the data sends the data to the first applicant.
\end{enumerate}
We illustrate the two last previous situations on the figure \ref{MPIEXCHANGE}.
\begin{figure}[h]
	\begin{center}
	    \psfrag{case 2}{\tiny case 2}
	    \psfrag{case 3}{\tiny case 3}
             \includegraphics[scale=0.35]{FIGURES/ComGroup.eps} 
	\end{center}
          \caption{On the left, we illustrate communications inside the same group. On the right, we illustrate the communications between  two MPI-groups. The dashed circle indicated the previous MPI group. In both cases, the masters process have the key 
          role to distribute informations for the exchange. The requests (\texttt{mpi\_isend} and \texttt{mpi\_irecv}) are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
\end{figure}

In the previous section, we reveled the importance of the data exchange.   the scheduler, its perfect execution is relevant to  the packet manager.
The packet manager  manage the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
between the MPI groups. The packet manager is unique for every MPI-group, it is based on a singleton pattern.

The role of the p

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN BIBLIO
\bibliographystyle{abbrv}
\bibliography{biblio}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END BIBLIO

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN PACKAGE
\documentclass[11pt]{amsart}       % AMS template just for the draft
\usepackage{geometry}                 % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                        % .letter or a4paper or a5paper or ... 
%\geometry{landscape}                 % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}     % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}                   % FIgure
\usepackage{amssymb}                 % Math symbol
\usepackage{epstopdf}                   % for working in .pdf and  not .dvi (convert .eps, .jpg,  etc to .pdf)
\usepackage{psfrag}
\usepackage[T1]{fontenc}
\usepackage{aeguill}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKAGE
\graphicspath{{PLOT/SinglePDGEMM/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS
\title{AMBIENT; an  execution and an MPI-CUDA scheduler  }
\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
%\author[UNIGE, ETHZ, CSCS]{UNIGE, ETHZ, CSCS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS

@@

\section{Design of the framework}
%In this section, we describe our framework in simple terms, without  a deep technical description.
%We will focus on the main functionalities  and show the essential features.

@@

Ambient employs an extension on the CUDA approach to a distributed cluster. This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which an runs an independent thread. In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc  \, must be considered.

The first thing coming in mind when we compare a  cluster and a GPU is the number of available cores.   Assimilate a single  CPU core  to  GPU core is reducer  because a CPU has additional capacities than a GPU core.
Thus, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$. Of course, the user has full control of the size and can adapt it to the treated problem. Like in CUDA, blocks (i.e. groups of "threads") and a grid of blocks can then be formed in Ambient. A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.

@@

\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Determination of the directed  acyclic graph   &    \\ \hline
                           \textbf{Require}    A class Matrix  that fits with Ambient framework   &  \\  
                           and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
                           \begin{tabular}{c l} 
                               \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j$  to the size of the vector \textbf{do} \\                               
                               \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
                               \tiny{4:}   & \hspace{0.8 cm}  $A[i]+ \!\! =B[j]$, $\vartriangleright$  push l/c kernel 1, \\
                               \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =3$,  $\vartriangleright$  push l/c kernel 2, \\                   
                               \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
                               \tiny{8:}   & \hspace{0.8 cm} $C[i]\times \!\! =D[j]$,  $\vartriangleright$ push l/c kernel 3, \\
                               \tiny{9:}   & \hspace{0.4 cm}  \textbf{end if} \\       
                               \tiny{10:} & \hspace{0.2 cm}  \textbf{end do} \\   
                               \tiny{11:} & \textbf{end do} \\                                 
                               \tiny{12:} & \texttt{playout()} $\vartriangleright$  execute the stack \\
                               \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
                                                & that push the good logistics and computational kernel. \\                               
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements. The association of a cumulative stack and  delate stack execution will determine the  directed  acyclic graph. \label{GRAPH}}
\end{table}

@@

When we execute a computational kernel, we get the data by the Ambient command  \texttt{current(\textit{A})(i,j)} (ref. \ref{CURRENT}), where \textit{A} is a matrix 
compatible with Ambient, and the indices $i$ and $j$ are the coordinates of the work group (Fig. \ref{MPIGROUP}). Three possibilities are possibles for getting the data when a process execute this functions
in a computational kernel.
\begin{enumerate}
\item The work group is already in the memory of the process. The access is direct.
\item The work group is in the memory space of the present MPI-group (previously created by the associated logistics kernel). The applicant process sends a query request (a specific packet) to its  master process of its MPI group.
The master process of the group gets the request (the master process knows the repartition of the workgroup inside the MPI-group), it sends a new request to
the actual owner, when the actual owner gets the request, it send the workgroup needed to the first applicant. 
\item The work group is inside an other MPI-group. As previously, the applicant process sends to the master process a query  request, and the master process of 
the actual MPI group  sends an other request to the older owner group. The older master process gets the message and sends a message to actual owner of the data. Then the owner 
of the data sends the data to the first applicant.
\end{enumerate}
We illustrate the two last previous situations on the figure \ref{MPIEXCHANGE}.
\begin{figure}[h]
	\begin{center}
	    \psfrag{case 2}{\tiny case 2}
	    \psfrag{case 3}{\tiny case 3}
             \includegraphics[scale=0.35]{FIGURES/ComGroup.eps} 
	\end{center}
          \caption{On the left, we illustrate communications inside the same group. On the right, we illustrate the communications between  two MPI-groups. The dashed circle indicated the previous MPI group. In both cases, the masters process have the key 
          role to distribute informations for the exchange. The requests  are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
\end{figure}

In the previous section, we reveled the importance of the communications, theirs perfect execution is relevant to  the packet manager.
The packet manager  manages the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
between the MPI groups. The packet manager is unique for every MPI-group, it is based on a singleton pattern.

 In the previous paragraph, we presented different communication blueprint when datas are requested by a process.
As communication path between the applicant  and the owner of the data can be more and less complex,  to solve every communication scheme,  we designed the packet manager on two main things :
the delegate pattern and a finite state machine (FSM).

To reach the best performance in HPC,  synchronization operations should be avoided  \cite{EXASCALE-2010}, because it is  a bottleneck.
Thus, we privileged  MPI asynchronous communication into Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Nevertheless, if a message 
is sent by a asynchronous instruction, an MPI test is necessary to check if the packet can be received.  As we refused synchronization instructions,
we repeat the process until receiving the message by an infinite loop. Thus we integrated the delegate pattern and the FSM inside the loop.

\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Algorithm of the packet manager   &    \\ \hline
                           \textbf{Require}  a MPI process require a block of data &  \\                           
                            \begin{tabular}{c l}                                
                               \tiny{1:}   & \textbf{for}  ;;  \textbf{do} $\vartriangleright$  infinite loop \\
                               \tiny{2:}   & \hspace{0.2 cm}  \texttt{Delegate()}  $\vartriangleright$ check if  packet has been, \\
                                                & and received and re-send if needed \\
                               \tiny{3:}   & \hspace{0.2 cm}  FSM $\vartriangleright$ break the loop, \\
                                               & if all packets arrived at destination. \\
                               \tiny{4:}   & \textbf{end do} \\
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Description of the packet manager}
\end{table}

\begin{figure}[h]
	\begin{center}
	    \psfrag{1}{\tiny \hspace{-0.1 cm}  1}
	    \psfrag{2}{\tiny \hspace{-0.1 cm}  2}
	    \psfrag{Loose}{\tiny \hspace{-0.2 cm}   Loose}
	    \psfrag{Closure}{\tiny \hspace{-0.2 cm}  Closure}
	    \psfrag{Open}{\tiny \hspace{-0.4 cm}  Opening}
             \includegraphics[scale=0.45]{FIGURES/FSM.eps} 
	\end{center}
          \caption{Schematic of the finite sate machine for the closing procedure. The states 1 and 2 correspond to open or close. The transitions are loose, closure or opening, they  depend of the situation of all packets (if they all reach theirs targets or not),
                          the loose transition means the packet has not been received, thus we iterate one more time on the infinite loop.}  \label{FSM}
\end{figure}

A delegate pattern (C\#) encapsulated methods,    It is similar to pointer of function under C++  but is allows a safer execution, we ported this pattern into Ambient.
The packet manager class contains a delegate object. The delegate class contains a series of operations pill-up  in a stack. These operations consists of forwarding requests, 
 block of data,  layout of the MPI group (block distribution) and theirs updates. As this delegate is included, into an infinite loop, the packets
will propagate process by process after every iterations (as shown on Fig. \ref{EXECUTION}). At the end of the delegate execution, a finite state machine checks the states of the packets. 
If all packets received (data packet, layout packet, \ldots), the FSM (Fig. \ref{FSM}) breaks the infinite loop.

@@

\begin{figure}
\include{PLOT/SinglePDGEMM/SinglePGEMM}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN BIBLIO
\bibliographystyle{abbrv}
\bibliography{biblio}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END BIBLIOr
